Precise Data Flow Analysis\\in the Presence of Correlated Method Calls

### 2015-03-05 oopsla2015/content/background.tex
rename folders

\subsection{IFDS}\label{sec:bgifds}
The IFDS framework~\cite{reps1995precise} is a precise and efficient algorithm for data-flow analysis
%. IFDS was developed in 1995 by T.\,Reps, S.\,Horwitz, and M.\,Sagiv at the University of Wisconsin and
that has been used to solve a variety of data-flow analysis problems~\cite{bodden2013spl,naeem2008typestate,DBLP:conf/birthday/KreikerRRSWY13,tripp2009taj}.
%\mtodo{ The four references take away space in the References section. Remove some of them?  }
The IFDS framework is an instance of the \textit{functional approach} to data-flow analysis~\cite{pnueli1981two}
because it constructs summaries of the effects of called procedures.
\commentout{The IFDS analysis is a version of the classic \textit{functional approach} to data-flow analysis proposed by M.\,Sharir and A.\,Pnueli~\cite{pnueli1981two}.}
\commentout{In other data-flow algorithms not based on the functional approach, the result of the analysis at the entry of a procedure ``merges'' the incoming data obtained from all callers of the procedure. As a consequence, there is one global data-flow result computed at the end of the procedure. Context-sensitivity, however, allows an analysis to compute the data-flow result for a given procedure as a \textit{function} of the data-flow value at the start of the procedure. 
In other words, the analysis result for a procedure varies depending on where the procedure was called from. This significantly improves the precision of a data-flow analysis, which is why context-sensitivity is an important advantage of IFDS over classic data-flow algorithms.}
The IFDS framework is applicable to \textit{interprocedural} data flow problems whose domain is \textit{subsets} of a \textit{finite} set $D$,
and whose data-flow functions are \textit{distributive}.
A function $f$ is distributive if 
$f(x_1\sqcap x_2)=f(x_1)\sqcap f(x_2)$.

@@

The input to the IFDS algorithm is specified as
%Formally, a data-flow analysis problem is suitable for an IFDS analysis if it can be encoded as an IFDS problem
  $(G^*,\,D,\,F,\,M_F,\,\sqcap)$,
where $G^*=(N^*,\,E^*)$ is the supergraph of the input program with nodes $N^*$ and edges $E^*$,
$D$ is a finite set of \textit{data-flow facts},
$F$ is a set of distributive dataflow functions of type $2^D\to2^D$,
$M_F:\,E^*\to F$ assigns a dataflow function to each supergraph edge,
and $\sqcap$ is the \textit{meet operator} on the powerset $2^D$, either union or intersection.
In our presentation, the meet operator will always be union, but all of the results apply dually when the
meet is intersection.

@@

\commentout{
is a function that maps supergraph edges to dataflow functions, and $M_F$ is extended to paths by composition\footnote{%
  Let $A$ be a set and $f:\,E^*\to(A\to A)$ a function from supergraph edges to functions on $A$.
  We say that $f$ is extended to  paths by composition to denote that for a path $q$ consisting of the edges $e_1,\,\dots,\,e_k$, $f(q)=f(e_k)\circ\ldots\circ f(e_1)\circ\id$.
}. The \textit{meet operator} $\sqcap$ is either union or intersection.
}

@@

\subsubsection{Overview of the IFDS Algorithm}\label{sec:overviewifds}
The key idea behind the IFDS algorithm is that it is possible to represent any distributive
function $f$ from $2^D$ to $2^D$ by a \textit{representation relation} $R_f \subseteq 
(D \cup \{0\})
\times
(D \cup \{0\})$. The representation relation can be visualized as a bipartite graph
with edges from one instance of $D \cup \{0\}$ to another instance of $D \cup \{0\}$.
The IFDS algorithm uses such graphs to efficiently represent both the input dataflow functions
and the summary functions that it computes for called procedures.
Specifically, the representation relation $R_f$ of a function $f$ is defined as:
\[
R_f=\{(\mathbf0,\,\mathbf0)\}\cup
      \{(\mathbf0,\,d_j)\,|\,d_j\in f(\varnothing)\}\cup
      \{(d_i,\,d_j)\,|\,d_j\in f(\{d_i\}) \setminus f(\varnothing)\}.
\]
%Each pair $(d_i,\,d_j)\in R_f$ corresponds to an edge $((n_1,\,d_i),\,(n_2,\,d_j))$ in the exploded supergraph.

\commentout{
Note that $R_f$ constructs pairs of dataflow facts so that
\begin{itemize}
  \item there is always an edge $(\mathbf0,\,\mathbf0)$ corresponding to the CFG edge;
  \item if there is an edge $(\mathbf0,\,d_j)$, then there is no other edge leading to $d_j$; in particular, there is never an edge $(d_i,\,\mathbf0)$ where $d_i\ne\mathbf0$.
\end{itemize}
}

\begin{example}\label{ex:flowfun}
    Given $D=\{u,\,v,\,w\}$ and $f(S) = S\setminus\{v\} \cup \{u\}$,
    the representation relation 
  %The representation relation $R_f$ for a set of data-flow facts 
  %and dataflow function $f=\lambda S\,.\,S\setminus\{v\}\cup\{u\}$ looks as follows:
  $
    R_f=\{(\mathbf0,\,\mathbf0),\,(\mathbf0,\,u),\,(w,\,w)\}
$, which can be visualized with the following graph:
   \begin{center}
    \tikzstyle{n}=[fill=black,circle,inner sep=1.5pt]
\begin{tikzpicture}[>=stealth,on grid, auto]
    \node [n,label=$\mathbf{0}$] (zero){};
    \node [n,] (zero_) [below=\edge of zero] {};
    \node [n,label=$u$] (u) [right=\nd of zero] {};
    \node [n,] (u_) [below=\edge of u] {};
    \node [n,label=$v$] (v) [right=\nd of u] {};
    \node [n,] (v_) [below=\edge of v] {};
    \node [n,label=$w$] (w) [right=\nd of v] {};
    \node [n,] (w_) [below=\edge of w] {};
    \path[->](zero) edge (zero_);
    \path[->](zero) edge (u_);
    \path[->](w) edge (w_);
  \end{tikzpicture}
  \end{center}
\end{example}

@@

Composition of two distributive
functions $f$ and $f'$ corresponds to finding reachable nodes in a graph composed from their representation
relations $R_f$ and $R_{f'}$. Therefore, evaluating the composed dataflow function for a control
flow path corresponds to finding reachable nodes in a graph composed from the representation relations
of the dataflow functions for individual instructions.

It is this graph of representation relations that the IFDS algorithm operates on.
In this graph, called the \textit{exploded supergraph}, each node is a pair $(n,d)$, where
$n\in N^*$ is a node of the control-flow supergraph and $d$ is an element of $D\cup\{0\}$.
For each edge $(n \to n') \in E^*$, the exploded supergraph contains a set of edges
$(n,d_i) \to (n',d_j)$, which form the representation relation of the dataflow function
$M_F(n \to n')$. The IFDS algorithm finds all exploded supergraph edges that are reachable by
\textit{realizable} paths in the exploded supergraph. A path
is \textit{realizable} if its projection to the (non-exploded) supergraph is a valid
path (i.e., if it is of the form $(n_0, d_0) \to (n_1,d_1) \to \cdots \to (n_m,d_m)$
and where $n_0 \to n_1 \to \cdots \to n_m$ is a valid path).

@@

\commentout{
A practical implementation of the IFDS algorithm generally takes as input a representation of the
exploded supergraph edges $E^\sharp$ instead of explicit dataflow functions $M_F$. This is convenient because
the two are equivalent in expressiveness, and because the IFDS algorithm works internally with the exploded
supergraph. More specifically, the input generally provides a function
$f: N^* \times D \times N^* \to 2^D$. Given a (non-exploded) supergraph edge $n \to n'$ and a dataflow fact $d$,
$f(n,d,n')$ returns the set of all $d'$ such that the exploded supergraph contains the edge $(n,d) \to (n',d')$.
For convenience, the function $f$ can be split up into separate functions that handle the cases when
the $n\to n'$ is an intraprocedural edge, a call-to-start edge, or an end-to-return edge.
%\mtodo{Do we leave the start-to-return edge out as a special case of intraprocedural edge?}
}

@@

The IDE algorithm~\cite{sagiv1996precise} extends IFDS to
\textit{interprocedural} \textit{distributive} \textit{environment}
problems. An \textit{environment} problem is an analysis whose dataflow
lattice is the lattice $\textsf{Env}(D,L)$ of maps from a finite set $D$ to a meet semilattice
$L$ of finite height, ordered pointwise. Like IFDS, IDE requires the dataflow functions to be distributive.

@@

The input to the IDE algorithm is
  $(G^*,\,D,\,L,\,M_\textsf{Env})$
where $G^*$ is a control-flow supergraph,
$D$ is a set of data-flow facts,
$L$ is a meet semilattice of finite height,
and $M_\textsf{Env}:\,E^*\to(\textsf{Env}(D,\,L)\to \textsf{Env}(D,\,L))$ assigns a dataflow function
to each supergraph edge.

@@

In order to do this efficiently, the IDE algorithm requires a representation of micro-functions
that is general enough to express the basic micro-functions of the dataflow functions for individual
instructions, and that supports computing the meet and composition of micro-functions.

%Similar to the IFDS algorithm,
A practical implementation of the IDE algorithm requires
the input dataflow functions to be provided in their pointwise representation
as exploded supergraph edges labelled with micro-functions.
%\mtodo{Should we cite the ``Practical extensions'' paper here that talks about the on-demand IFDS algorithm?}
Specifically,
the input is generally provided as a function $\textsf{EdgeFn}: (N^* \times D) \times (N^* \times D) \to F$,
where $F$ is the set of representations of micro-functions from $L$ to $L$.
Given an exploded supergraph edge $e=(n,\,d) \to (n',\,d')$,
$\textsf{EdgeFn}((n,\,d),\,(n',\,d'))$ returns the micro-function that appears on the exploded supergraph edge $e$.
%Like in the IFDS algorithm,
In an implementation, it can be convenient to split the function $\textsf{EdgeFn}$ into separate functions that handle the cases when $n\to n'$ is an intraprocedural edge, a call-to-return edge, a call-to-start edge, or an end-to-return edge.

>>>>

\subsection{IFDS}\label{sec:bgifds}
The IFDS framework~\cite{reps1995precise} is a precise and efficient algorithm for data-flow analysis
%. IFDS was developed in 1995 by T.\,Reps, S.\,Horwitz, and M.\,Sagiv at the University of Wisconsin and
that has been used to solve a variety of data-flow analysis problems~\cite{bodden2013spl,naeem2008typestate,DBLP:conf/birthday/KreikerRRSWY13,tripp2009taj}.
%\mtodo{ The four references take away space in the References section. Remove some of them?  }
The IFDS framework is an instance of the \textit{functional approach} to data-flow analysis~\cite{pnueli1981two}
because it constructs summaries of the effects of called procedures.
\commentout{The IFDS analysis is a version of the classic \textit{functional approach} to data-flow analysis proposed by M.\,Sharir and A.\,Pnueli~\cite{pnueli1981two}.}
\commentout{In other data-flow algorithms not based on the functional approach, the result of the analysis at the entry of a procedure ``merges'' the incoming data obtained from all callers of the procedure. As a consequence, there is one global data-flow result computed at the end of the procedure. Context-sensitivity, however, allows an analysis to compute the data-flow result for a given procedure as a \textit{function} of the data-flow value at the start of the procedure. 
In other words, the analysis result for a procedure varies depending on where the procedure was called from. This significantly improves the precision of a data-flow analysis, which is why context-sensitivity is an important advantage of IFDS over classic data-flow algorithms.}
The IFDS framework is applicable to \textit{interprocedural} data-flow problems whose domain is \textit{subsets} of a \textit{finite} set $D$,
and whose data-flow functions are \textit{distributive}.
A function $f$ is distributive if 
$f(x_1\sqcap x_2)=f(x_1)\sqcap f(x_2)$.

@@

The input to the IFDS algorithm is specified as
%Formally, a data-flow analysis problem is suitable for an IFDS analysis if it can be encoded as an IFDS problem
  $(G^*,\,D,\,F,\,M_F,\,\sqcap)$,
where $G^*=(N^*,\,E^*)$ is the supergraph of the input program with nodes $N^*$ and edges $E^*$,
$D$ is a finite set of \textit{data-flow facts},
$F$ is a set of distributive data-flow functions of type $2^D\to2^D$,
$M_F:\,E^*\to F$ assigns a data-flow function to each supergraph edge,
and $\sqcap$ is the \textit{meet operator} on the powerset $2^D$, either union or intersection.
In our presentation, the meet operator will always be union, but all of the results apply dually when the
meet is intersection.

@@

\commentout{
is a function that maps supergraph edges to data-flow functions, and $M_F$ is extended to paths by composition\footnote{%
  Let $A$ be a set and $f:\,E^*\to(A\to A)$ a function from supergraph edges to functions on $A$.
  We say that $f$ is extended to  paths by composition to denote that for a path $q$ consisting of the edges $e_1,\,\dots,\,e_k$, $f(q)=f(e_k)\circ\ldots\circ f(e_1)\circ\id$.
}. The \textit{meet operator} $\sqcap$ is either union or intersection.
}

@@

\subsubsection{Overview of the IFDS Algorithm}\label{sec:overviewifds}
The key idea behind the IFDS algorithm is that it is possible to represent any distributive
function $f$ from $2^D$ to $2^D$ by a \textit{representation relation} $R_f \subseteq 
(D \cup \{0\})
\times
(D \cup \{0\})$. The representation relation can be visualized as a bipartite graph
with edges from one instance of $D \cup \{0\}$ to another instance of $D \cup \{0\}$.
The IFDS algorithm uses such graphs to efficiently represent both the input data-flow functions
and the summary functions that it computes for called procedures.
Specifically, the representation relation $R_f$ of a function $f$ is defined as:
\[
R_f=\{(\mathbf0,\,\mathbf0)\}\cup
      \{(\mathbf0,\,d_j)\,|\,d_j\in f(\varnothing)\}\cup
      \{(d_i,\,d_j)\,|\,d_j\in f(\{d_i\}) \setminus f(\varnothing)\}.
\]
%Each pair $(d_i,\,d_j)\in R_f$ corresponds to an edge $((n_1,\,d_i),\,(n_2,\,d_j))$ in the exploded supergraph.

\commentout{
Note that $R_f$ constructs pairs of data-flow facts so that
\begin{itemize}
  \item there is always an edge $(\mathbf0,\,\mathbf0)$ corresponding to the CFG edge;
  \item if there is an edge $(\mathbf0,\,d_j)$, then there is no other edge leading to $d_j$; in particular, there is never an edge $(d_i,\,\mathbf0)$ where $d_i\ne\mathbf0$.
\end{itemize}
}

\begin{example}\label{ex:flowfun}
    Given $D=\{u,\,v,\,w\}$ and $f(S) = S\setminus\{v\} \cup \{u\}$,
    the representation relation 
  %The representation relation $R_f$ for a set of data-flow facts 
  %and data-flow function $f=\lambda S\,.\,S\setminus\{v\}\cup\{u\}$ looks as follows:
  $
    R_f=\{(\mathbf0,\,\mathbf0),\,(\mathbf0,\,u),\,(w,\,w)\}
$, which can be visualized with the following graph:
   \begin{center}
    \tikzstyle{n}=[fill=black,circle,inner sep=1.5pt]
\begin{tikzpicture}[>=stealth,on grid, auto]
    \node [n,label=$\mathbf{0}$] (zero){};
    \node [n,] (zero_) [below=\edge of zero] {};
    \node [n,label=$u$] (u) [right=\nd of zero] {};
    \node [n,] (u_) [below=\edge of u] {};
    \node [n,label=$v$] (v) [right=\nd of u] {};
    \node [n,] (v_) [below=\edge of v] {};
    \node [n,label=$w$] (w) [right=\nd of v] {};
    \node [n,] (w_) [below=\edge of w] {};
    \path[->](zero) edge (zero_);
    \path[->](zero) edge (u_);
    \path[->](w) edge (w_);
  \end{tikzpicture}
  \end{center}
\end{example}

@@

Composition of two distributive
functions $f$ and $f'$ corresponds to finding reachable nodes in a graph composed from their representation
relations $R_f$ and $R_{f'}$. Therefore, evaluating the composed data-flow function for a control
flow path corresponds to finding reachable nodes in a graph composed from the representation relations
of the data-flow functions for individual instructions.

It is this graph of representation relations that the IFDS algorithm operates on.
In this graph, called the \textit{exploded supergraph}, each node is a pair $(n,d)$, where
$n\in N^*$ is a node of the control-flow supergraph and $d$ is an element of $D\cup\{0\}$.
For each edge $(n \to n') \in E^*$, the exploded supergraph contains a set of edges
$(n,d_i) \to (n',d_j)$, which form the representation relation of the data-flow function
$M_F(n \to n')$. The IFDS algorithm finds all exploded supergraph edges that are reachable by
\textit{realizable} paths in the exploded supergraph. A path
is \textit{realizable} if its projection to the (non-exploded) supergraph is a valid
path (i.e., if it is of the form $(n_0, d_0) \to (n_1,d_1) \to \cdots \to (n_m,d_m)$
and where $n_0 \to n_1 \to \cdots \to n_m$ is a valid path).

@@

\commentout{
A practical implementation of the IFDS algorithm generally takes as input a representation of the
exploded supergraph edges $E^\sharp$ instead of explicit data-flow functions $M_F$. This is convenient because
the two are equivalent in expressiveness, and because the IFDS algorithm works internally with the exploded
supergraph. More specifically, the input generally provides a function
$f: N^* \times D \times N^* \to 2^D$. Given a (non-exploded) supergraph edge $n \to n'$ and a data-flow fact $d$,
$f(n,d,n')$ returns the set of all $d'$ such that the exploded supergraph contains the edge $(n,d) \to (n',d')$.
For convenience, the function $f$ can be split up into separate functions that handle the cases when
the $n\to n'$ is an intraprocedural edge, a call-to-start edge, or an end-to-return edge.
%\mtodo{Do we leave the start-to-return edge out as a special case of intraprocedural edge?}
}

@@

The IDE algorithm~\cite{sagiv1996precise} extends IFDS to
\textit{interprocedural} \textit{distributive} \textit{environment}
problems. An \textit{environment} problem is an analysis whose data-flow
lattice is the lattice $\textsf{Env}(D,L)$ of maps from a finite set $D$ to a meet semilattice
$L$ of finite height, ordered pointwise. Like IFDS, IDE requires the data-flow functions to be distributive.

@@

The input to the IDE algorithm is
  $(G^*,\,D,\,L,\,M_\textsf{Env})$
where $G^*$ is a control-flow supergraph,
$D$ is a set of data-flow facts,
$L$ is a meet semilattice of finite height,
and $M_\textsf{Env}:\,E^*\to(\textsf{Env}(D,\,L)\to \textsf{Env}(D,\,L))$ assigns a data-flow function
to each supergraph edge.

@@

In order to do this efficiently, the IDE algorithm requires a representation of micro-functions
that is general enough to express the basic micro-functions of the data-flow functions for individual
instructions, and that supports computing the meet and composition of micro-functions.

%Similar to the IFDS algorithm,
A practical implementation of the IDE algorithm requires
the input data-flow functions to be provided in their pointwise representation
as exploded supergraph edges labelled with micro-functions.
%\mtodo{Should we cite the ``Practical extensions'' paper here that talks about the on-demand IFDS algorithm?}
Specifically,
the input is generally provided as a function $\textsf{EdgeFn}: (N^* \times D) \times (N^* \times D) \to F$,
where $F$ is the set of representations of micro-functions from $L$ to $L$.
Given an exploded supergraph edge $e=(n,\,d) \to (n',\,d')$,
$\textsf{EdgeFn}((n,\,d),\,(n',\,d'))$ returns the micro-function that appears on the exploded supergraph edge $e$.
%Like in the IFDS algorithm,
In an implementation, it can be convenient to split the function $\textsf{EdgeFn}$ into separate functions that handle the cases when $n\to n'$ is an intraprocedural edge, a call-to-return edge, a call-to-start edge, or an end-to-return edge.
